---
title: "Stock Price Forecasting Using 30-Second Time Bucket Intervals"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(rugarch)
library(ggplot2)
library(gridExtra)
library(xgboost)

stock <- read.csv("C:/Users/Ame PC/Desktop/3rd year 1st sem/DATA 3888/Data/stock_1.csv")
```

#EDA
```{r}
colSums(is.na(stock))
str(stock)
head(stock)

```
#Compute Statistics 
```{r}
stock = stock%>%
  mutate(WAP = (bid_price1 * ask_size1 + ask_price1 * bid_size1) / (bid_size1 + ask_size1), BidAskSpread = ask_price1 / bid_price1 - 1)

head(stock)


#log return

Time_IDs = unique(stock$time_id)[1:500]

log_r_stock <- list()

for (i in 1:length(Time_IDs)) {
  seconds = stock %>% filter(time_id == Time_IDs[i]) %>%
    pull(seconds_in_bucket)
  price = stock %>% filter(time_id == Time_IDs[i]) %>% 
    pull(WAP)
  log_r = log(price[-1] / price[1:length(price) - 1]) 
  log_r_stock[[i]] = tibble(time = seconds[-1], log_return = log_r)
  
  no_diff <- (1:600)[!(1:600 %in% log_r_stock[[i]]$time)]
  if (length(no_diff) >0){
    no_diff_df = tibble(time = no_diff, log_return = 0)
    log_r_stock[[i]] = log_r_stock[[i]] %>% 
      bind_rows(no_diff_df) %>%
      arrange(time)
  }
    
    
}

head(log_r_stock[[1]])
head(log_r_stock[[500]])


vol = list()
comp_vol <- function(x) sqrt(sum(x^2))

for (i in 1:500) {
  log_r_stock[[i]] = log_r_stock[[i]] %>% 
    mutate(time_bucket = ceiling(time/30))
  vol[[i]] = log_r_stock[[i]] %>% 
    group_by(time_bucket) %>% 
    summarise(volatility = comp_vol(log_return))
}

head(vol[[1]])
```


## ACF
```{r}
acf(log_r_stock[[1]]$log_return, main = "ACF graph for log returns")

acf(vol[[1]]$volatility, main = "ACF graph for realised volatility")
```


# Model Building

## Model 1
```{r}
#Weighted linear regression
train_vol = list()
valid_vol = list()

for (i in 1:500) {
  train_vol[[i]] = vol[[i]][1:16, ]
  valid_vol[[i]] = vol[[i]][-(1:16), ]
  
  
}


predict = stock%>% 
  mutate( num_order = bid_size1 + ask_size1 + bid_size2 + ask_size2, time_bucket = ceiling(seconds_in_bucket / 30)) %>% 
  select(time_id, time_bucket, WAP, num_order, BidAskSpread)

list_reg <- list()
for (i in 1:500) {
  list_reg[[i]] <- predict%>% 
    filter(time_id == Time_IDs[i], time_bucket != 0) %>%
    group_by(time_bucket) %>%
    summarise(price = mean(WAP), 
              order = mean(num_order), 
              BidAskSpread = mean(BidAskSpread)) %>% 
    slice(1:15) %>%  
    mutate(volatility = train_vol[[i]]$volatility[-1])
}
weight_model =lm(volatility ~ price + order + BidAskSpread, data = list_reg[[1]], weights = 0.8 ^ seq(from = 7, to = 0, by = -0.5))

summary(weight_model)
  
weight_models = list()
for (i in 1:500){
  weight_models[[i]] = lm(volatility ~ price + order + BidAskSpread, data = list_reg[[i]], weights = 0.8 ^ seq(from = 7, to = 0, by = -0.5))

}

summary(weight_models[[200]])


#Prediction 
val_list_reg <- list()
predict_lm <- list()
for (i in 1:500) { 
  val_list_reg[[i]] = predict %>% 
    filter(time_id == Time_IDs[i], time_bucket != 0) %>% 
      group_by(time_bucket) %>% 
    summarise(price = mean(WAP), order = mean(num_order), BidAskSpread = mean(BidAskSpread)) %>% 
    slice(16:19)
  
  predict_lm[[i]] = predict(weight_models[[i]], newdata = val_list_reg[[i]])
}

predict_lm[[1]]

print(data.frame(
  actual    = valid_vol[[1]]$volatility,
  predicted = predict_lm[[1]]
))



#summary of stats

MSE_wlm <- numeric(500)
QLIKE_wlm <- numeric(500)

for (i in 1:500) { 
  actual <- valid_vol[[i]]$volatility 
  pred <- predict_lm[[i]]
  pred[pred <= 0] <- 1e-10
    
  MSE_wlm[i] <- mean((actual - pred)^2)
  QLIKE_wlm[i] <- mean(actual/pred - log(actual/pred) - 1)
  
}
  
  summary_wlm <- data.frame(
    
    Metrics_to_evaluate = c("MSE", "QLIKE"), 
    Mean = c(mean(MSE_wlm, na.rm= TRUE), mean(QLIKE_wlm, na.rm = TRUE)),
    Median = c(median(MSE_wlm, na.rm= TRUE), median(QLIKE_wlm, na.rm = TRUE)), 
    SD = c(sd(MSE_wlm, na.rm= TRUE), sd(QLIKE_wlm, na.rm = TRUE)), 
    Min = c(min(MSE_wlm, na.rm= TRUE), min(QLIKE_wlm, na.rm = TRUE)),
    Max = c(max(MSE_wlm, na.rm= TRUE), max(QLIKE_wlm, na.rm = TRUE)))
    
    
    
   print(summary_wlm)
    
  


metrics_df <- data.frame(
  MSE   = MSE_wlm,
  QLIKE = QLIKE_wlm
)

plot_mse <- ggplot(metrics_df, aes(y = MSE)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  scale_y_log10() +
  labs(title = "Weighted Linear Regression: MSE",
       y     = "MSE") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

plot_qlike <- ggplot(metrics_df, aes(y = QLIKE)) +
  geom_boxplot(fill = "coral", alpha = 0.7) +
  scale_y_log10() +
  labs(title = "Weighted Linear Regression: QLIKE",
       y     = "QLIKE") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

plot_mse
plot_qlike


co_ef_df <- data.frame(
  intercept    = sapply(weight_models, function(m) coef(m)[1]),
  price        = sapply(weight_models, function(m) coef(m)[2]),
  order        = sapply(weight_models, function(m) coef(m)[3]),
  BidAskSpread = sapply(weight_models, function(m) coef(m)[4])
)


summary(co_ef_df)


cat("Average equation across all Time_IDs:")
cat("Volatility =", round(mean(co_ef_df$intercept), 6),
    "+", round(mean(co_ef_df$price), 6), "* price",
    "+", round(mean(co_ef_df$order), 9), "* order",
    "+", round(mean(co_ef_df$BidAskSpread), 6), "* BidAskSpread")


```

## Model 2 
```{r}
#Arma-Garch

spec = ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 1)), 
                  mean.model = list(armaOrder = c(1, 1)), 
                  distribution.model = "norm")

ARMA_GARCH_models = list()
for (i in 1:500) {
  ARMA_GARCH_models[[i]] = ugarchfit(spec = spec, 
                                     data = log_r_stock[[i]] %>% 
                                       filter(time <= 480) %>% 
                                       pull(log_return),
                                      solver = "hybrid")
}

Real_vol_pred = rep(0, length(vol))

set.seed(111)
for(i in 1:500) {
  a_g_spec <- getspec(ARMA_GARCH_models[[i]])
  setfixed(a_g_spec) <- as.list(coef(ARMA_GARCH_models[[i]]))
  future_path = fitted(ugarchpath(a_g_spec, n.sim = 30, m.sim = 1000))
  
  val_columns <- colSums(is.na(future_path)) <= 5
  val_paths <- future_path[, val_columns]
  
  if (ncol(val_paths) >= 100){
    
    row_medians <- apply(val_paths, 1, median, na.rm = TRUE)
    for (j in 1:nrow(val_paths)) {
      val_paths[j, is.na(val_paths[j,])] <- row_medians[j]
    }
    Real_vol_pred[i] <- mean(sqrt(colSums(val_paths^2)))
    
    }
  else{
    Real_vol_pred[i] <- NA
  }
}
  
  
cat("Successfully predicted: ", sum(!is.na(Real_vol_pred)), "/ 500\n")
cat("Path unrealiable: ", sum(is.na(Real_vol_pred)), " / 500")


real_vol_per_path <- sqrt(colSums(future_path^2, na.rm = TRUE))

ggplot(data.frame(rv = real_vol_per_path), aes(x = rv)) + geom_histogram(fill = "goldenrod", color = "white", bins = 50, alpha = 0.7) + geom_vline(xintercept = mean(real_vol_per_path), color = "pink", linewidth = 1, linetype = "dotted") + annotate("text", x = mean(real_vol_per_path) * 1.05, y = 60, label = paste("Predicted Realised Volatility = ", round(mean(real_vol_per_path), 6)), color = "red") + labs(title = "Distribution of Predicted Volatility", subtitle = "Based on ARMA GARCH Model 1000 simulations", x = "Realised Volatility", y = "Count") + theme_minimal()



MSE_garch   <- numeric(500)
QLIKE_garch <- numeric(500)

for (i in 1:500) {
  
  if (is.na(Real_vol_pred[i])) next
  
  actual <- valid_vol[[i]]$volatility
  

  pred   <- rep(Real_vol_pred[i], 4)
  pred[pred <= 0] <- 1e-10
  
  MSE_garch[i]   <- mean((actual - pred)^2)
  QLIKE_garch[i] <- mean(actual/pred - log(actual/pred) - 1)
}

sum_garch <- data.frame(
  Metric = c("MSE", "QLIKE"),
  Median = c(median(MSE_garch,   na.rm = TRUE),
             median(QLIKE_garch, na.rm = TRUE)),
  Mean   = c(mean(MSE_garch,   na.rm = TRUE),
             mean(QLIKE_garch, na.rm = TRUE)),
  SD     = c(sd(MSE_garch,   na.rm = TRUE),
             sd(QLIKE_garch, na.rm = TRUE)),
  Min    = c(min(MSE_garch,   na.rm = TRUE),
             min(QLIKE_garch, na.rm = TRUE)),
  Max    = c(max(MSE_garch,   na.rm = TRUE),
             max(QLIKE_garch, na.rm = TRUE))
)

cat("ARMA-GARCH RESULTS")
print(sum_garch)

arm_plot_1 <- ggplot(
  data.frame(val = MSE_garch[MSE_garch > 0]),
  aes(y = val)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  scale_y_log10() +
  labs(title = "ARMA-GARCH: MSE",
       y     = "MSE") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

arm_plot_2 <- ggplot(
  data.frame(val = QLIKE_garch[QLIKE_garch > 0 & 
                                is.finite(QLIKE_garch)]),
  aes(y = val)) +
  geom_boxplot(fill = "coral", alpha = 0.7) +
  scale_y_log10() +
  labs(title = "ARMA-GARCH: QLIKE",
       y     = "QLIKE") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

arm_plot_1
arm_plot_2

print(sum_garch)

```
only 1 path is unrealiable > as more than 5 within that path were NA > 

From the graph > Our ARMA-GARCH model simulated 1000 possible futures for the next 30 seconds. The most likely realised volatility is around 0.001644. However the distribution shows meaningful uncertainty — there's a right tail suggesting occasional volatility spikes you should account for when pricing options.

pink line > final averaged predicted rv line 



## Model 3
```{r}
#Xgboost model 

set.seed(111)


valid_index <- sapply(1:500, function(i) {
  micro <- predict %>%
    filter(time_id == Time_IDs[i],
           time_bucket != 0) %>%
    group_by(time_bucket) %>%
    summarise(spread = mean(BidAskSpread),
              .groups = "drop")
  nrow(micro) >= 20
})



rows_train <- list()
rows_test <- list()


for(i in 1:500){
  
  if(!valid_index[i]) next
  
  real_vol <- train_vol[[i]]$volatility
  micro <- predict %>%
    filter(time_id == Time_IDs[i], time_bucket != 0)%>% 
    group_by(time_bucket) %>% 
    summarise(
      spread = mean(BidAskSpread),
      .groups = "drop")

cat("Valid time ids", sum(valid_index))  
  
  for(t in 6:16){
    rows_train[[length(rows_train) + 1]] <- data.frame(
      real_vol_lag1 = real_vol[t-1],
      real_vol_lag2 = real_vol[t-2],
      mean_real_vol_5 = mean(real_vol[(t-5):(t-1)]), 
      sd_real_vol_5 = sd(real_vol[(t-5):(t-1)]), 
      spread = micro$spread[t], 
      target = real_vol[t],
      index = i 
    )
  }
  
  for(t in 17:20){
  rows_test[[length(rows_test) + 1]] <- data.frame( 
    real_vol_lag1 = real_vol[t-1],
      real_vol_lag2 = real_vol[t-2],
      mean_real_vol_5 = mean(real_vol[(t-5):(t-1)]), 
      sd_real_vol_5 = sd(real_vol[(t-5):(t-1)]), 
      spread = micro$spread[t],  
      target = NA_real_, 
      index = i
    
    )    
  }
}
  
   train_df <- bind_rows(rows_train) 
   test_df <- bind_rows(rows_test)
    
   
   feature  <- c("real_vol_lag1","real_vol_lag2","mean_real_vol_5",
            "sd_real_vol_5","spread")
d_train <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature]),
  label = train_df$target
)
d_test <- xgb.DMatrix(
  data = as.matrix(test_df[, feature])
)
   
  grid_param <- expand.grid(
  eta       = c(0.05, 0.1),
  max_depth = c(3, 4)
) 
  
  cat("Hyperparameter Tunings: ")
  
  cross_val_res <- data.frame(
  eta = numeric(nrow(grid_param)),
  max_depth    = numeric(nrow(grid_param)),
  best_nrounds = numeric(nrow(grid_param)),
  cv_rmse      = numeric(nrow(grid_param)))
  
  
  for(z in 1:nrow(grid_param)) { 
    cv <- xgb.cv(
      params = list(
        objective = "reg:squarederror", 
        eta = grid_param$eta[z], 
       max_depth        = grid_param$max_depth[z],
      subsample        = 0.8,
      colsample_bytree = 0.8
    ),
    data                  = d_train,
    nrounds               = 500,
    nfold                 = 5,
    early_stopping_rounds = 20,
    verbose               = 0,
    metrics               = "rmse"
  )
  
     best_n <- ifelse(
    is.null(cv$best_iteration) || length(cv$best_iteration) == 0,
    nrow(cv$evaluation_log),
    cv$best_iteration
  )
    
  cross_val_res[z, ] <- data.frame(
    eta          = grid_param$eta[z],
    max_depth    = grid_param$max_depth[z],
    best_nrounds = best_n, 
    cv_rmse      = min(cv$evaluation_log$test_rmse_mean)
  )
  
  cat("eta:", grid_param$eta[z],
      "| depth:", grid_param$max_depth[z],
      "| trees:", cross_val_res$best_nrounds[z],
      "| RMSE:", round(cross_val_res$cv_rmse[z], 8), "\n")
}
    
    best_res <- cross_val_res[which.min(cross_val_res$cv_rmse), ]
cat("BEST PARAMETERS")
print(best_res)
    

xgb_fin_md <- xgb.train( 
  param = list(
    objective        = "reg:squarederror",
    eta              = best_res$eta,
    max_depth        = best_res$max_depth,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data    = d_train,
  nrounds = best_res$best_nrounds,
  verbose = 0
)

cat("Final model trained:")
cat("  eta =      ", best_res$eta,        "\n")
cat("  max_depth =", best_res$max_depth,  "\n")
cat("  nrounds =  ", best_res$best_nrounds, "\n")
  
  
import <- xgb.importance( 
  feature_names = feature, 
  model = xgb_fin_md) 

cat("FEATURE IMPORTANCE")
print(import)

xgb.plot.importance(
  import,
  main = "XGBoost: Feature Importance",
  col  = "forestgreen"
)

#test

test_df$pred <- pmax(predict(xgb_fin_md, d_test), 0)

pred_xgb <- lapply(1:500, function(i) {
  test_df %>%
    filter(index == i) %>%
    pull(pred)
})


cat("Actual vs Predicted (time_id 1):")
print(data.frame(
  actual    = valid_vol[[1]]$volatility,
  predicted = pred_xgb[[1]]
))

#plot graph 

MSE_xgb  <- numeric(500)
QLIKE_xgb <- numeric(500)

for (i in 1:500) {
  actual <- valid_vol[[i]]$volatility
  pred   <- pred_xgb[[i]]
  
  if (length(pred) == 0 || any(is.na(pred))) next
  
  pred[pred <= 0] <- 1e-10
  MSE_xgb[i]     <- mean((actual - pred)^2)
  QLIKE_xgb[i]   <- mean(actual/pred - log(actual/pred) - 1)
}

sum_xgb <- data.frame(
  Metric = c("MSE", "QLIKE"),
  Median = c(median(MSE_xgb,   na.rm = TRUE),
             median(QLIKE_xgb, na.rm = TRUE)),
  Mean   = c(mean(MSE_xgb,   na.rm = TRUE),
             mean(QLIKE_xgb, na.rm = TRUE)),
  SD     = c(sd(MSE_xgb,   na.rm = TRUE),
             sd(QLIKE_xgb, na.rm = TRUE))
)

cat("XGBOOST RESULTS")
print(sum_xgb)

  
plot_1 <- ggplot(
  data.frame(val = MSE_xgb[MSE_xgb > 0]),
  aes(y = val)) +
  geom_boxplot(fill = "pink", alpha = 0.8) +
  scale_y_log10() +
  labs(title = "Evaluate XG Boost: MSE",
       y     = "MSE") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

plot_2 <- ggplot(
  data.frame(val = QLIKE_xgb[QLIKE_xgb > 0]),
  aes(y = val)) +
  geom_boxplot(fill = "coral", alpha = 0.7) +
  scale_y_log10() +
  labs(title = "Evaluate XG Boost: QLIKE",
       y     = "QLIKE") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
  
plot_1 
plot_2

```
machine learning model 



## Evaluation
```{r}
final_comparison_table <- data.frame(
  Model        = c("WLS Regression",
                   "ARMA GARCH",
                   "XG Boost"),
  Median_MSE   = c(median(MSE_wlm,   na.rm = TRUE),
                   median(MSE_garch, na.rm = TRUE),
                   median(MSE_xgb,   na.rm = TRUE)),
  Median_QLIKE = c(median(QLIKE_wlm[is.finite(QLIKE_wlm)],
                          na.rm = TRUE),
                   median(QLIKE_garch[is.finite(QLIKE_garch)],
                          na.rm = TRUE),
                   median(QLIKE_xgb[is.finite(QLIKE_xgb)],
                          na.rm = TRUE)),
P95_QLIKE    = c(quantile(QLIKE_wlm[is.finite(QLIKE_wlm)],   0.95, na.rm = TRUE),
                   quantile(QLIKE_garch[is.finite(QLIKE_garch)], 0.95, na.rm = TRUE),
                   quantile(QLIKE_xgb[is.finite(QLIKE_xgb)],    0.95, na.rm = TRUE))
) %>% arrange(P95_QLIKE)

print(final_comparison_table)


all_mse <- data.frame(
  MSE = c(
    MSE_wlm[MSE_wlm > 0 & is.finite(MSE_wlm)],
    MSE_garch[MSE_garch > 0 & is.finite(MSE_garch)],
    MSE_xgb[MSE_xgb > 0 & is.finite(MSE_xgb)]
  ),
  Model = c(
    rep("WLS Regression",
        sum(MSE_wlm > 0 & is.finite(MSE_wlm))),
    rep("ARMA GARCH",
        sum(MSE_garch > 0 & is.finite(MSE_garch))),
    rep("XG Boost",
        sum(MSE_xgb > 0 & is.finite(MSE_xgb)))
  )
) %>%
  mutate(Model = factor(
    Model,
    levels = c("WLS Regression",
               "ARMA GARCH",
               "XG Boost")
  ))

plot_mse_all <- ggplot(
  all_mse,
  aes(x = Model, y = MSE, fill = Model)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  scale_y_log10() +
  scale_fill_manual(
    values = c("steelblue", "pink", "plum")
  ) +
  labs(
    title    = "Model Comparison Using MSE",
    subtitle = "Lower means better prediction accuracy",
    x        = "",
    y        = "MSE"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

plot_mse_all

all_qlike <- data.frame(
  QLIKE = c(
    QLIKE_wlm[QLIKE_wlm > 0 & is.finite(QLIKE_wlm)],
    QLIKE_garch[QLIKE_garch > 0 & is.finite(QLIKE_garch)],
    QLIKE_xgb[QLIKE_xgb > 0 & is.finite(QLIKE_xgb)]
  ),
  Model = c(
    rep("WLS Regression",
        sum(QLIKE_wlm > 0 & is.finite(QLIKE_wlm))),
    rep("ARMA GARCH",
        sum(QLIKE_garch > 0 & is.finite(QLIKE_garch))),
    rep("XG Boost",
        sum(QLIKE_xgb > 0 & is.finite(QLIKE_xgb)))
  )
) %>%
  mutate(Model = factor(
    Model,
    levels = c("WLS Regression",
               "ARMA GARCH",
               "XG Boost")
  ))

plot_qlike_all <- ggplot(
  all_qlike,
  aes(x = Model, y = QLIKE, fill = Model)) +
  geom_boxplot(alpha = 0.8) +
  scale_y_log10() +
  scale_fill_manual(
    values = c("palegreen", "peachpuff", "paleturquoise")
  ) +
  labs(
    title    = "Model Comparison Using QLIKE",
    subtitle = "Lower means the model is better at penalising",
    x        = "",
    y        = "QLIKE"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

plot_qlike_all 


best_model <- final_comparison_table$Model[1]
cat("Getting prediction for:", best_model)

best_pred <- switch(best_model,
  "WLS Regression" = predict_lm,
  "ARMA GARCH"     = lapply(1:500, function(i) 
                      rep(Real_vol_pred[i], 4)),
  "XG Boost"        = pred_xgb
)


best_time <- which.min(sapply(1:500, function(i) {
  pred <- best_pred[[i]]
  if (length(pred) == 0 || any(is.na(pred))) return(Inf)
  mean((valid_vol[[i]]$volatility - pred)^2)
}))

plot_df <- data.frame(
  bucket = vol[[best_time]]$time_bucket,
  actual = vol[[best_time]]$volatility,
  pred   = c(rep(NA, 16), best_pred[[best_time]])
)

ggplot(plot_df, aes(x = bucket)) +
  geom_line(aes(y = actual),
            colour   = "blue",
            linewidth = 0.8) +
  geom_point(aes(y = actual),
             colour = "blue",
             size   = 1.5) +
  geom_point(aes(y = pred),
             colour  = "orchid",
             size    = 3,
             na.rm   = TRUE) +
  geom_vline(xintercept = 16.5,
             linetype   = "dashed",
             colour     = "grey") +
  annotate("text",
           x     = 13,
           y     = max(plot_df$actual, na.rm = TRUE),
           label = "Training",
           colour = "grey",
           size  = 3.5) +
  annotate("text",
           x     = 18.5,
           y     = max(plot_df$actual, na.rm = TRUE),
           label = "Validation",
           colour = "orchid",
           size  = 3.5) +
  labs(
    title    = paste(best_model,
                     ": Actual vs Predicted of best example"),
    subtitle = "Black = actual | Red = predicted",
    x        = "30 second bucket",
    y        = "Realised Volatility"
  ) +
  theme_minimal()

```


